Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-29 text/ms.tex
Edit discussion

We found strong evidence that the multiplicative process noise jumps from time
step to time step were greater than we would expect under a normal distribution
and a Gompertz or Ricker-logistic model in \overallMinPerc--\overallMaxPerc\%
of populations assessed. We detected black swans more frequently for longer
time series and for populations with a smaller magnitude of process error, as
we might expect. However, we failed to find strong evidence that black swans
were associated with density dependence, population growth rate, or lifespan.
In cases that we verified, black swans were often a result of the interaction
between elements of extreme climate, predators and parasite cycles, and strong
changes in human pressures. Our empirical results, sensitivity tests, and
simulation tests suggest that estimating the tail shape of process error is
a viable method of detecting heavy-tailed population dynamics and if anything
will underestimate the probability of black swans. The presence of black swans
highlights the importance of developing management strategies that detect
quickly, respond to, and are robust to extremes in population dynamics ---
particularly as the frequency and magnitude of climatic extremes increase over
the next century \citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. Previous work with an older
version of the GPDD assessed the distribution of abundance time series but
focused on identifying if the lognormal distribution was the most frequently
parsimonious model \citep{halley2002}. For heavy-tailed distributions,
\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
four-parameter Levy stable distribution and found little information criteria
support for these distributions in longer time series. However, black swan
events are by definition rare, and the majority of time series in the GPDD are
short for these purposes. Therefore, we would not expect to observe black swans
in a large proportion of populations. Furthermore, by probabilistically
quantifying process noise tail behaviour on a continuous spectrum, and allowing
for non-stationary time series and density dependence, our analysis allows for
a more nuanced description of the evidence of ecological black swans. In an
earlier study, \citet{keitt1998} described heavy (power law) tails in breeding
bird population abundance. However, this finding was challenged by
\citet{allen2001}, who showed that mixing data across species could create
falsely generate heavy tails. This rebuttal supports our use of the
t distribution to represent heavy tails, since a t distribution can be
represented by a mixture of normal distributions \citep[with
inverse-gamma-distributed variances,][]{gelman2014}.

@@

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point. Recording
mistakes undoubtedly comprise some of the black swans we detected. Conversely,
recording mistakes could mask heavy tails if initial observations are assumed
to be recording errors and are discarded or altered. Third, the temporal scales
of observation and population dynamics vary considerably across populations in
the GPDD and this likely influences the detection of heavy tails. For example,
if we make frequent observations relative to generation time (e.g.~for many
large-bodied mammals) we will average across generations and perhaps miss
black swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans do occur, can we forecast the probability of
black swans in space and time? Furthermore, what management policies allow us
to detect them quickly after they happen? Secondly, can we isolate the
components of ecological dynamics that experience black swans events by moving
from phenomenological models such as the Gompertz to mechanistic models, that
take into account, for example, recruitment? We expect that greater insight
into the mechanisms and covariates of ecological black swans may be best
obtained through specific geographic and taxonomic subsets of data where longer
time series are available \citep[e.g.][]{segura2013}. Finally, what is the
impact of allowing for black swans in forecasts of ecological risk? Financial
analysts are concerned with the shape of financial-return distribution tails
because these impact estimates of risk --- the probability of a specific
magnitude of undesired event occurring \citep{rachev2008}. It is possible that
a comparable focus in ecology would alter our perception of ecological
forecasts such as extinction risk.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans. A similar focus could benefit resource management once our ability to
predict the spatial probability and covariates of ecological black swans
improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by extreme disease or environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and `unexpected',
`surprising', or `counterintuitive' ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and tune future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
extreme climate \citep{meehl2004,ipcc2012}.

@@

The following supplementary material is available online for this article:\\
Figures S1--S11\\
Table S1\\
Example Stan code for a heavy-tailed Gompertz model and multilevel beta
regression\\
The GPDD IDs used in our analysis\\
All code and data to recreate our analysis is available at:\\
\url{https://github.com/seananderson/heavy-tails}

>>>>

We found strong evidence that the process noise in ecological time series was heavy-tailed under the Gompertz or Ricker-logistic models in \overallMinPerc--\overallMaxPerc\%
of populations assessed. We detected black swans more frequently for longer
time series and for populations with a smaller magnitude of process error, as
we might expect. However, we failed to find strong evidence that black swans
were associated with density dependence, population growth rate, or lifespan.
In cases that we verified, black swans were often a result of the interaction
between elements of extreme climate, predators and parasite cycles, and strong
changes in human pressures. Our empirical results, sensitivity analyses, and
simulation tests suggest that estimating the tail shape of process noise is
a viable method of detecting black swan population dynamics and if anything
will underestimate the probability of black swans. The presence of black swans
highlights the importance of developing management strategies that detect
quickly, respond to, and are robust to extremes in population dynamics ---
particularly as the frequency and magnitude of climatic extremes increase over
the next century \citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. Previous work with an older
version of the GPDD assessed the distribution of abundance time series but
focused on identifying if the lognormal distribution was the most frequently
parsimonious model \citep{halley2002}. For heavy-tailed distributions,
\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
four-parameter Levy stable distribution and found little information criteria
support for these distributions in longer time series. However, black swan
events are by definition rare, and the majority of time series in the GPDD are
short for these purposes. Therefore, we would not expect to observe black swans
in a large proportion of populations. Furthermore, by probabilistically
quantifying process noise tails on a continuous spectrum, and allowing
for non-stationary time series and density dependence, our analysis allows for
a more nuanced description of the evidence of ecological black swans. In an
earlier study, \citet{keitt1998} described heavy (power law) tails in breeding
bird population abundance. However, this finding was challenged by
\citet{allen2001}, who showed that mixing data across species could create
falsely generate heavy tails. This rebuttal supports our use of the
t distribution to represent heavy tails, since a t distribution can be
represented by a mixture of normal distributions \citep[with
inverse-gamma-distributed variances,][]{gelman2014}.

@@

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point;
it is likely that some of the black swans in our dataset are 
recording mistakes mistakes.
Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed errors. Third, the temporal scales
of observation and population dynamics vary considerably across populations in
the GPDD and this likely influences the detection of heavy tails. For example,
if we make frequent observations relative to generation time (e.g.~for many
large-bodied mammals) we will average across generations and perhaps miss
black swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans do occur, can we forecast the probability of
black swans in space and time? Furthermore, what management policies allow us
to detect them quickly after they happen? Can we isolate the
components of ecological dynamics that experience black swans events by moving
from phenomenological models such as the Gompertz to mechanistic models, that
take into account, for example, recruitment? We expect that greater insight
into the mechanisms and covariates of ecological black swans may be best
obtained through specific geographic and taxonomic subsets of data where longer
time series are available \citep[e.g.][]{segura2013}. Finally, what is the
impact of allowing for black swans in forecasts of ecological risk? Financial
analysts are concerned with the shape of financial-return-distribution tails
because these impact estimates of risk --- the probability of a specific
magnitude of undesired event occurring \citep{rachev2008}. It is possible that
a comparable focus in ecology would alter our perception of ecological
forecasts such as extinction risk.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF).
A similar focus could benefit resource management once our ability to
predict the spatial probability and covariates of ecological black swans
improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and tune future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
climate extremes \citep{meehl2004,ipcc2012}.

@@

The following supplementary material is available online for this article:\\
Figures S1--S11\\
Table S1\\
Example Stan code for a heavy-tailed Gompertz model and multilevel beta
regression\\
The GPDD IDs used in our analysis\\
All code and data to recreate our analysis are available at:\\
\url{https://github.com/seananderson/heavy-tails}
